### 2023-02-07 chapter1/chapter1.tex
Added SVG files to repository
Various minor re-wordings and corrections
Change list:
Made use of past tense consistent in methods
Removed unnecessary transitions like "also" and "then," especially in the methods

\begin{abstract}
\noindent
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.
\end{abstract}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{graphical_overview.png}
\centering
\caption{\textbf{Overview of homology inference pipeline.}
\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation. Sequences that share homology are colored with shades of the same hue. \textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
\label{fig:graphical_overview}
\end{figure}

\section*{Results}
\subsection*{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).

@@

Though there are benchmarks available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

@@

\section*{Materials and methods}
\subsection*{Sequence de-duplication and BLAST search parameters}
The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene. Thus, the first accession associated with a sequence and gene pair is the sequence’s representative accession for the gene. BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}. An E-value cutoff of 1 was used for the initial searches. However, this cutoff was lowered to 1E-10 during processing of the BLAST output.

\subsection*{Extraction of HSPs from BLAST output}
To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene. The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP. Iterating over the groups, all the HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group. This method collects all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeds the highest-scoring HSP of the next best gene. This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins. If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they do not overlap with any other HSP previously marked as disjoint. Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were then filtered by overlap and reciprocity criteria. The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs. This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences. The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.

\subsection*{Clustering by \textit{k}-clique percolation}
\textit{k}-clique percolation was implemented in two steps. In the first, maximal cliques were identified. In the second, a percolation graph is constructed by defining edges between cliques if they have \textit{k}-1 nodes in common. Clusters are the connected components of this second graph. The first step uses the NetworkX implementation of a maximal clique algorithm. The second step uses a modification of the NetworkX implementation of the \textit{k}-clique community algorithm. The NetworkX implementation exhaustively finds all edges in the percolation graph. Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach is needlessly expensive for large graphs. The custom implementation instead uses a progressive approach where each clique is checked against a list of known communities, merging communities as necessary in each step.

The hit graph is sparse, so these algorithms are efficient when applied to its individual connected components. However, some components have a structure with many maximal cliques, which dramatically slows the first or second step of the clique percolation algorithm. Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead. Out of over 10,000 connected components, only seven timed out, and many of those contained highly dense clusters of histone sequences.

\subsection*{Addition of paralogs to orthologous groups}
The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches. The resulting output was processed identically except the HSPs were not filtered using the best gene criterion. Thus, all HSPs for each query were merged into hits. The best hit for each query and target gene was then chosen as the hit with the highest sum of bit scores from disjoint HSPs. (Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits which exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs. The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}. Representative sequences for each gene were then selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were then grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was then scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence is aligned, the profile incentivizes the selection of sequences contain the fewest gaps and match the consensus alignment. Unfortunately, this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, so the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

\subsection*{Alignment refinement}
The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7. A binary profile was created from the alignment using Gaussian process sequence weighting as described in the section on selecting representative sequences. (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5. The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10. Diverged regions were taken as the complement of the highly conserved regions. For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4. The resulting sub-alignments were then stitched into the initial alignment.

\subsection*{Alignment curation}
The alignments were coded into gap and non-gap symbols to simplify the emission distributions. The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text. The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively. The first component was a two-state Markov process which was in turn composed of two subprocess. The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips. The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns. The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM. However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data. Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree would correspond to a single sequence in the alignment.

@@

The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM. Regions with a high probability of state 2 or 3 are candidates for trimming. However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changes most abruptly. Instead, the following algorithm was used. First, a high cutoff defines a ``seed'' region. The left and right endpoints of the seed are then expanded both inwards and outwards to define two intervals from which boundaries are selected. The outward expansion halts when the probability or its derivative is below two distinct thresholds, respectively. The inward expansion halts when the derivative is below a different threshold. The left and right boundaries are chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns. The gap profile is calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences. By combining where the model’s confidence changes rapidly with the observed change in the gap pattern, this method generally selects reasonable boundaries.

 States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first. Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method. First, the probability of state 3 of columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions. Regions were then defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively. The outer and inner derivative cutoffs were both 0.001. The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step. This ensured that any regions which were intermediate between state 2 and 3 were included. Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.75. The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method. However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05. These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.

\subsection*{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Then orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were then merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).

To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations. Some protein sequences are ``low quality,'' meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes. The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple ``reverse translation'' of the alignment. After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained. Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.

To fit trees using the two-state GTR model of substitution, the protein alignments were first coded into gap or non-gap symbols. As before, 100 meta-alignments were constructed from these coded alignments for each sampling strategy. In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments. Trees were then fit using the GTR2 model with no rate categories. An invariant category, however, was included if invariant columns were allowed. Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were then merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).

>>>>

\begin{abstract}
\noindent
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.
\end{abstract}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{graphical_overview/out/graphical_overview.png}
\centering
\caption{\textbf{Overview of homology inference pipeline.}
\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation. Sequences that share homology are colored with shades of the same hue. \textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
\label{fig:graphical_overview}
\end{figure}

\section*{Results}
\subsection*{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).

@@

Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

@@

\section*{Materials and methods}
\subsection*{Sequence de-duplication and BLAST search parameters}
The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene. Thus, the first accession associated with a sequence and gene pair was the sequence’s representative accession for the gene. BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}. An E-value cutoff of 1 was used for the initial searches. However, this cutoff was lowered to 1E-10 during processing of the BLAST output.

\subsection*{Extraction of HSPs from BLAST output}
To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene. The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP. Iterating over the groups, all HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group. This method collected all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeded the highest-scoring HSP of the next best gene. This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins. If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint. Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were filtered by overlap and reciprocity criteria. The overlap criterion was applied first and required that 50\% of residues in the query were aligned in compatible HSPs. This excluded false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits spanned a sufficient fraction of the query and target sequences. The reciprocity criterion required each query-target pair had a corresponding hit where the roles were reversed, which ensured there was no ambiguity in which target was the best match for the query.

\subsection*{Clustering by \textit{k}-clique percolation}
\textit{k}-clique percolation was implemented in two steps. In the first, maximal cliques were identified. In the second, a percolation graph was constructed by defining edges between cliques if they had \textit{k}-1 nodes in common. Clusters were the connected components of this second graph. The first step used the NetworkX implementation of a maximal clique algorithm. The second step used a modification of the NetworkX implementation of the \textit{k}-clique community algorithm. The NetworkX implementation exhaustively finds all edges in the percolation graph. Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach was needlessly expensive for large graphs. The custom implementation instead used a progressive approach where each clique was checked against a list of known communities, merging communities as necessary in each step.

The hit graph was sparse, so these algorithms were efficient when applied to its individual connected components. However, some components had a structure with many maximal cliques, which dramatically slowed the first or second step of the clique percolation algorithm. Thus, if either step exceeded 90 s, the process timed out, and the simpler \textit{k}-core algorithm was used instead. Out of over 10,000 connected components, only seven timed out, and many of those contained highly dense clusters of histone sequences.

\subsection*{Addition of paralogs to orthologous groups}
The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches. The resulting output was processed identically except the HSPs were not filtered using the best gene criterion. Thus, all HSPs for each query were merged into hits. The best hit for each query and target gene was chosen as the hit with the highest sum of bit scores from disjoint HSPs. (Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs. The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}. Representative sequences for each gene were selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence was aligned, the profile prioritized the selection of sequences with the fewest gaps that best matched the consensus alignment. Since this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

\subsection*{Alignment refinement}
The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7. A binary profile was created from the alignment using Gaussian process sequence weighting as described in the section on selecting representative sequences. (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5. The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10. Diverged regions were taken as the complement of the highly conserved regions. For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4. The resulting sub-alignments were stitched into the initial alignment.

\subsection*{Alignment curation}
The alignments were coded into gap and non-gap symbols to simplify the emission distributions. The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text. The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively. The first component was a two-state Markov process which was in turn composed of two subprocess. The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips. The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns. The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM. However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data. Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree corresponded to a single sequence in the alignment.

@@

The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM. Regions with a high probability of state 2 or 3 were candidates for trimming. However, because the probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries of these regions as the columns where the gap pattern changed most abruptly. Instead, the following algorithm was used. First, a high cutoff defined a ``seed'' region. The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected. The outward expansion halted when the probability or its derivative was below two distinct thresholds, respectively. The inward expansion halted when the derivative was below a different threshold. The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns. The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.

 States 2 and 3 have distinct characteristics which required different trimming strategies. Regions with a high probability of state 3 were handled first. Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they would not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns using the following method. First, the probability of state 3 for columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions. Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively. The outer and inner derivative cutoffs were both 0.001. The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step. This ensured that any regions which were intermediate between state 2 and 3 were included. Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.75. The posterior probabilities from the missing data phylo-HMM were converted into state assignments using a similar method. However, the initial seeds were defined with a cutoff of 0.75, and the seeds were expanded outward to the first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.05. These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.

\subsection*{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model and four discrete gamma rate categories using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}).

To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations. Some protein sequences were ``low quality,'' meaning their coding sequences contained frameshifts, premature stop codons, or other errors even though they were strong hits to known protein-coding genes. The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple ``reverse translation'' of the alignment. After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained. Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.

To fit trees using the two-state GTR model of substitution, the protein alignments were first coded into gap or non-gap symbols. As before, 100 meta-alignments were constructed from these coded alignments for each sampling strategy. In this case, only the presence of invariant columns was varied, yielding two sets of meta-alignments. Trees were fit using the GTR2 model with no rate categories. An invariant category, however, was included if invariant columns were allowed. Since the bootstrap confidences were sometimes lower than 50\%, the resulting trees from each set were merged into a loose consensus tree to prevent multifurcations (Fig.~\ref{sfig:trees_GTR}).
